<!-- R Commander Markdown Template -->

Replace with Main Title
=======================

### chard

### `r as.character(Sys.Date())`

```{r echo=FALSE}
# include this code chunk as-is to set options
knitr::opts_chunk$set(comment=NA, prompt=TRUE, out.width=750, fig.height=8, fig.width=8)
library(Rcmdr)
library(car)
library(RcmdrMisc)
```

```{r}
library(abind, pos=18)
```


```{r}
library(e1071, pos=19)
```



### Numerical Summaries: EAWE01
```{r}
numSummary(EAWE01[,"EARNINGS", drop=FALSE], statistics=c("mean", "sd", "skewness", 
  "kurtosis"), quantiles=c(0,.25,.5,.75,1), type="2")
```


### Numerical Summaries: EAWE01
```{r}
numSummary(EAWE01[,"EARNINGS", drop=FALSE], statistics=c("mean", "sd", "skewness", 
  "kurtosis"), quantiles=c(0,.25,.5,.75,1), type="2")
```


### Numerical Summaries: EAWE01
```{r}
numSummary(EAWE01[,"EARNINGS", drop=FALSE], statistics=c("mean", "sd", "skewness", 
  "kurtosis"), quantiles=c(0,.25,.5,.75,1), type="2")
```


### Boxplot: ~ EARNINGS
```{r}
Boxplot( ~ EARNINGS, data=EAWE01, id=list(method="y"))
```


### Normality Test: ~EARNINGS
```{r}
normalityTest(~EARNINGS, test="shapiro.test", data=EAWE01)
```


### Normality Test: ~EARNINGS
```{r}
normalityTest(~EARNINGS, test="shapiro.test", data=EAWE01)
```


### Correlation Test: EARNINGS, S
```{r}
with(EAWE01, cor.test(EARNINGS, S, alternative="two.sided", method="pearson"))
```


### Linear Model: RegModel.1: EARNINGS~S
```{r}
RegModel.1 <- lm(EARNINGS~S, data=EAWE01)
summary(RegModel.1)
```

<!-- R Commander Markdown Template -->

Replace with Main Title
=======================

### chard

### `r as.character(Sys.Date())`

```{r echo=FALSE}
# include this code chunk as-is to set options
knitr::opts_chunk$set(comment=NA, prompt=TRUE, out.width=750, fig.height=8, fig.width=8)
library(Rcmdr)
library(car)
library(RcmdrMisc)
```


### Scatterplot: Residuals...101~S
```{r}
scatterplot(Residuals...101~S, regLine=TRUE, smooth=FALSE, 
  id=list(method='mahal', n=2), boxplots=FALSE, data=EAWE01)
```


### Scatterplot: Residuals...101~S
```{r}
scatterplot(Residuals...101~S, regLine=TRUE, smooth=FALSE, 
  id=list(method='mahal', n=2), boxplots=FALSE, data=EAWE01)
```

```{r}
library(zoo, pos=20)
```


```{r}
library(lmtest, pos=20)
```



### Breusch-Pagan Test: EARNINGS ~ S
```{r}
bptest(EARNINGS ~ S, varformula = ~ fitted.values(RegModel.1), 
  studentize=FALSE, data=EAWE01)
```


### Histogram: Residuals...101
```{r}
with(EAWE01, Hist(Residuals...101, scale="percent", breaks="Sturges", 
  col="darkgray"))
```


### Boxplot: ~ Residuals...101
```{r}
Boxplot( ~ Residuals...101, data=EAWE01, id=list(method="y"))
```


### Boxplot: ~ Residuals...101
```{r}
Boxplot( ~ Residuals...101, data=EAWE01, id=list(method="y"))
```


### Normality Test: ~Residuals...101
```{r}
normalityTest(~Residuals...101, test="shapiro.test", data=EAWE01)
```


### RESET Test: EARNINGS ~ S
```{r}
resettest(EARNINGS ~ S, power=2:3, type="regressor", data=EAWE01)
```


### Linear Model: RegModel.2: EARNINGS~EXP+S
```{r}
RegModel.2 <- lm(EARNINGS~EXP+S, data=EAWE01)
summary(RegModel.2)
```


### Variance-Inflation Factors: RegModel.2
```{r}
vif(RegModel.2)
```


```{r}
round(cov2cor(vcov(RegModel.2)), 3) # Correlations of parameter estimates
```


### Akaike Information Criterion (AIC): RegModel.1
```{r}
AIC(RegModel.1)
```



### Bayesian Information Criterion (BIC): RegModel.1
```{r}
BIC(RegModel.1)
```


### Akaike Information Criterion (AIC): RegModel.2
```{r}
AIC(RegModel.2)
```



### Bayesian Information Criterion (BIC): RegModel.2
```{r}
BIC(RegModel.2)
```

<!-- R Commander Markdown Template -->

Replace with Main Title
=======================

### chard

### `r as.character(Sys.Date())`

```{r echo=FALSE}
# include this code chunk as-is to set options
knitr::opts_chunk$set(comment=NA, prompt=TRUE, out.width=750, fig.height=8, fig.width=8)
library(Rcmdr)
library(car)
library(RcmdrMisc)
```


### Scatterplot Matrix: ~EXP+Residuals+S
```{r}
scatterplotMatrix(~EXP+Residuals+S, regLine=TRUE, smooth=FALSE, 
  diagonal=list(method="density"), data=EAWE01)
```


### Breusch-Pagan Test: EARNINGS ~ EXP + S
```{r}
bptest(EARNINGS ~ EXP + S, varformula = ~ fitted.values(RegModel.2), 
  studentize=FALSE, data=EAWE01)
```


### Histogram: ResidualsModel2
```{r}
with(EAWE01, Hist(ResidualsModel2, scale="frequency", breaks="Sturges", 
  col="darkgray"))
```


### RESET Test: EARNINGS ~ EXP + S
```{r}
resettest(EARNINGS ~ EXP + S, power=2:3, type="regressor", data=EAWE01)
```


