---
title: "Multivar"
output: html_notebook
---

===============================================

Execute whole chunk: *Ctrl+Shift+Enter*
Execute Line within chunk: *Ctrl+Enter*
Create new chunk: *Ctrl+Alt+I*
Preview HTML output: *Ctrl+Shift+K*



```{r}
print('Correlation Numerical Test for variables EARNINGS and S')
  
```

===============================================


```{r, Load pkg}
library(pacman)
pacman::p_load(safetensors,reticulate,odbc,tensorflow,tidyverse,quantmod,psych,pastecs,olsrr,stargazer,ggfortify,caTools,car,MASS,corrplot,performance,flexdashboard,tidymodels,dplyr,ggplot2,tidyr,sparklyr,purrrlyr,GGally,openxlsx)
#R.version
```

===============================================

## Formula and Calculation of Multiple Linear Regression 
##https://rmd4sci.njtierney.com/math
##LaTeX
$$
\begin{aligned}&y_i = \beta_0 + \beta _1 x_{i1} + \beta _2 x_{i2} + \beta _3 x_{i3} +  ... + \beta _n x_{in} + \epsilon_i

\\&\textbf{where, for } i = n~\textbf{observations:}
\\&y_i=\text{dependent variable}
\\&x_i=\text{explanatory variables}
\\&\beta_0=\text{y-intercept (constant term)}
\\&\beta_n=\text{slope coefficients for each explanatory variable}
\\&\epsilon_i=\text{the model's error term (also known as the residuals)}
\end{aligned}
$$

===============================================



```{r, DataFrame}
df = read.xlsx("EAWE04.xlsx", sheet = 1)
```


```{r, EDA}
colnames(df)
view(df)

```

```{r, Correlation Test}
cor.test(df$EARNINGS, df$S)
```
===============================================

```{r}
##Script used in PCA class

EAWE01 <- within(EAWE01, {
  URBAN <- factor(URBAN, labels=c('Urban','Rural','Others'))
})
Boxplot(EARNINGS ~ URBAN, data=EAWE01, id=list(method="y"))
library(mvtnorm, pos=18)
library(survival, pos=18)
library(MASS, pos=18)
library(TH.data, pos=18)
library(multcomp, pos=18)
library(abind, pos=23)
AnovaModel.1 <- aov(EARNINGS ~ URBAN, data = EAWE01)
summary(AnovaModel.1)
with(EAWE01, numSummary(EARNINGS, groups = URBAN, statistics=c('mean', 'sd')))
AnovaModel.2 <- aov(EARNINGS ~ URBAN, data = EAWE01)
summary(AnovaModel.2)
with(EAWE01, numSummary(EARNINGS, groups = URBAN, statistics=c('mean', 'sd')))
local({
  .Pairs <- glht(AnovaModel.2, linfct = mcp(URBAN = "Tukey"))
  print(summary(.Pairs)) # pairwise tests
  print(confint(.Pairs, level=0.95)) # confidence intervals
  print(cld(.Pairs, level=0.05)) # compact letter display
  old.oma <- par(oma=c(0, 5, 0, 0))
  plot(confint(.Pairs))
  par(old.oma)
})
LinearModel.3 <- lm(EARNINGS ~ URBAN*GENDER, data = EAWE01)
summary(LinearModel.3)
Anova(LinearModel.3, type="II")
with(EAWE01, plotMeans(EARNINGS, URBAN, error.bars="se", connect=TRUE))
local({
  .Table <- xtabs(~GENDER+URBAN, data=EAWE01)
  cat("\nFrequency table:\n")
  print(.Table)
  .Test <- chisq.test(.Table, correct=FALSE)
  print(.Test)
})




```




```{r}
## same as above but in markdown

```


<!-- R Commander Markdown Template -->

Replace with Main Title
=======================

### chard

### `r as.character(Sys.Date())`

```{r echo=FALSE}
# include this code chunk as-is to set options
knitr::opts_chunk$set(comment=NA, prompt=TRUE, out.width=750, fig.height=8, fig.width=8)
library(Rcmdr)
library(car)
library(RcmdrMisc)
```

```{r}
EAWE01 <- within(EAWE01, {
  URBAN <- factor(URBAN, labels=c('Urban','Rural','Others'))
})
```


### Boxplot: EARNINGS ~ URBAN
```{r}
Boxplot(EARNINGS ~ URBAN, data=EAWE01, id=list(method="y"))
```

```{r}
library(mvtnorm, pos=18)
```


```{r}
library(survival, pos=18)
```


```{r}
library(MASS, pos=18)
```


```{r}
library(TH.data, pos=18)
```


```{r}
library(multcomp, pos=18)
```


```{r}
library(abind, pos=23)
```



### One-Way Analysis of Variance: EARNINGS ~ URBAN
```{r}
AnovaModel.1 <- aov(EARNINGS ~ URBAN, data = EAWE01)
summary(AnovaModel.1)
with(EAWE01, numSummary(EARNINGS, groups = URBAN, statistics=c('mean', 'sd')))
```


### One-Way Analysis of Variance: EARNINGS ~ URBAN
```{r}
AnovaModel.2 <- aov(EARNINGS ~ URBAN, data = EAWE01)
summary(AnovaModel.2)
with(EAWE01, numSummary(EARNINGS, groups = URBAN, statistics=c('mean', 'sd')))
local({
  .Pairs <- glht(AnovaModel.2, linfct = mcp(URBAN = "Tukey"))
  print(summary(.Pairs)) # pairwise tests
  print(confint(.Pairs, level=0.95)) # confidence intervals
  print(cld(.Pairs, level=0.05)) # compact letter display
  old.oma <- par(oma=c(0, 5, 0, 0))
  plot(confint(.Pairs))
  par(old.oma)
})
```


### Linear Model: LinearModel.3: EARNINGS ~ URBAN*GENDER
```{r}
LinearModel.3 <- lm(EARNINGS ~ URBAN*GENDER, data = EAWE01)
summary(LinearModel.3)
```


### Analysis of Variance: LinearModel.3
```{r}
Anova(LinearModel.3, type="II")
```


### Plot of Means: EARNINGS by URBAN
```{r}
with(EAWE01, plotMeans(EARNINGS, URBAN, error.bars="se", connect=TRUE))
```


### Two-Way Contingency Table: GENDER, URBAN
```{r}
local({
  .Table <- xtabs(~GENDER+URBAN, data=EAWE01)
  cat("\nFrequency table:\n")
  print(.Table)
  .Test <- chisq.test(.Table, correct=FALSE)
  print(.Test)
})
```


===============================================





https://guides.nyu.edu/quant/choose_test_NDV



