---
title: "QuantDL_HW1"
author: Richard Gan 
date: "`r Sys.Date()`"
output:html
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>
# ===============================================
<------------- PKG ------------->
```{r,warning = FALSE,message = FALSE,pkg}
#install.packages('pacman')
require(pacman)
pacman::p_load(stats,RSQLite,tinytex,safetensors,reticulate,odbc,tensorflow,tidyverse,quantmod,psych,pastecs,olsrr,stargazer,ggfortify,caTools,car,MASS,corrplot,performance,flexdashboard,tidymodels,dplyr,ggplot2,tidyr,sparklyr,purrrlyr,GGally,openxlsx,YieldCurve,quantmod,odbc,getPass,quantreg,quantdr,quantdl)
```

<br>
# ===============================================
<------------- DESC ------------->

->>>> quantmod
Overview:
quantmod stands for "Quantitative Financial Modelling Framework" and is a popular package for financial modeling, trading, and data visualization in R.

Key Features:
Data Retrieval: It can download financial data from various sources (e.g., Yahoo Finance, Google Finance).
Charting: It provides tools for creating complex financial charts.
Technical Indicators: It includes a wide array of technical analysis indicators (e.g., moving averages, Bollinger Bands).
Data Manipulation: It supports data manipulation and transformation operations.

->>>> quantdr
Overview:
quantdr is less common and focuses on dimension reduction techniques for high-dimensional data, which is particularly useful in statistical modeling and machine learning.

Key Features:
Dimension Reduction: Implements various methods for dimension reduction such as Principal Component Analysis (PCA), Independent Component Analysis (ICA), and Factor Analysis.
Statistical Analysis: Provides tools for the statistical analysis of reduced dimension data.
Visualization: Offers visualization tools for high-dimensional data and the results of dimension reduction.

->>>> quantdl
Overview:
As of now, there isn't a well-known quantdl package in the R ecosystem. It's possible you might be referring to a hypothetical or specific-use package that isn't widely recognized or has been confused with another package.

Summary of Differences:
Purpose:

quantmod: Primarily focused on financial data retrieval, charting, and technical analysis.
quantdr: Focused on dimension reduction techniques for statistical and machine learning applications.
quantdl: No widely known package by this name, may be a typographical error or niche package.
Key Use Cases:

quantmod: Ideal for financial analysts, traders, and quantitative researchers working with market data.
quantdr: Suitable for data scientists and statisticians working with high-dimensional datasets requiring dimension reduction.

<br>
# ===============================================
<------------- CODE ------------->

```{r,quantdl}

getSymbols("AAPL", from = "2020-01-01", to = "2023-12-31")
head(AAPL)
chartSeries(AAPL, theme = chartTheme("white"), name = "Apple Inc. (AAPL)")
addSMA(n = 50, col = "blue")
addSMA(n = 200, col = "red")
returns <- dailyReturn(AAPL)
chartSeries(returns, theme = chartTheme("white"), name = "Daily Returns of AAPL")
summary(returns)
# Install and load the package
install.packages("quantmod")
library(quantmod)

# import historical NSDQ:APPL
getSymbols("AAPL", from = "2020-01-01", to = "2023-12-31")

#View first rows
head(AAPL)

#Plot 
chartSeries(AAPL, theme = chartTheme("white"), name = "Apple Inc. (AAPL)")

# Add  indicators: 50-day and 200-day MA
addSMA(n = 50, col = "blue")
addSMA(n = 200, col = "red")

#compute + plot daily
returns <- dailyReturn(AAPL)
chartSeries(returns, theme = chartTheme("white"), name = "Daily Returns of AAPL")


summary(returns)


```
<br>
# ===============================================
<------------- END ------------->
